AERO 590: Comparing Hydrogen and Jet-A for a Ultra High-Bypass Turbofan with Water Recirculation

### 2022-12-21 AE590_report/main.tex
updated report

\documentclass[12pt]{article}
% \documentclass[12pt]{new-aiaa}

@@

Water recirculation is the process of extracting water from the exhaust stream of a propulsion system and injecting it upstream of the combustor as finely atomized droplets.
NASA, Boeing, and Rolls-Royce studied injectiong water into the airstream and suggested that this technique reduces the NOx emissions as much as 47 percent~\cite{nasa_inject}.
Additionally, water recirculation improves fuel efficiency and thrust output with lower burner temperatures that can improve the lifetime of turbine blades and reduce noise~\cite{nasa_inject}.
Traditional propulsion systems that burn hydrocarbon fuels would require external water storage on the aircraft because they do not produce enough in the exhaust for recirculation.
The added weight of tanks, pumping, and ducting makes this concept infeasible for a conventional aircraft.
However, if hydrogen fuels are used, the main product of hydrogen combustion is water vapor and can thus be recovered.
The ability to recirculate water vapor from the exhaust of hydrogen combustion reduces the requirement for storage tanks and allows for the creation of a closed loop system inside the propulsion cycle.
Additionally, Pratt and Whitney are currently looking at closed-loop water vapor recovery for impementation in their Hydrogen Steam-Injected, Inter‐Cooled Turbine Engine (HySIITE) as part of the US Departement of Energy's Advanced Research Projects Agency-Energy (ARPA-E) program (CITE).
Pratt and Whitney suggest that this type of hydrogen-powered engine could achieve zero-carbon emissions and reduce NOx emissions by as much as 80\%.

@@

\noindent
To model this complicated water recirculation loop in pyCycle new components had to be developed.
The two new components that were developed are a water injector to add water to the flow and a water extractor to divert a fraction of the water in the flow back upstream.
The injector component is based on the combustor component already included in pyCycle which operates by injecting a given fuel-to-air ratio (FAR) or fuel flow rate to determine the mass of fuel added to the incoming flow.
This new mixture combination is then used to compute the various chemical species present in the flow at the given thermodynamic state which is determined by the incoming flow.
The new species composition and thermodynamic variables are then determined using a Gibbs Free Energy minization approach called Chemical Equilibrium with Applications (CEA).
Once the new composition is determined, the various thermodynamic variables are passed to the next component in the engine.
The water injector component was created to work similarly to the combustor but would instead inject water as a reactant instead of the fuel.
The water injector can take either water-to-air ratio (WAR) or water mass flow rate to determine how much water is added to the flow.
A simple schematic of the injector is shown in Figure \ref{fig:injector}

@@

\noindent
The water extractor component is based on an air-bleed component in pyCycle with the added complexity of extracting a fraction of a specific species within the flow.
This extractor component works by first determining the water content of the incoming flow and extracting a given fraction of that water flow.
The water content of the incoming flow is determined by solving a CEA analysis of the incoming flow.
This water content in the incoming flow is then used with a given fraction of the water to extract to determine the how much mass flow to remove from the stream.
At the some time, the atomic mixture is updated to represent the mixture that would be left if a number of hydrogen and oxygen atoms corresponding to the amount of water were removed.
The outputs of the extractor component are the flow rate of the main flow and the corresponding properties in addition to a water mass flow rate.
While this process certainly would have pressure loss penalties in terms of condensing water in a condersor, this project looks to just look at the effects of water recovery since there are no current models for nozzle exhaust condensors.
A simple schematic of the extractor is shown in Figure \ref{fig:extractor}

@@

\subsection{NOx Model}
To compute the Nitrogen Oxide (NOx) emissions from the engine model when using JetA as a fuel, the P3T3 NOx correlation was used (CITE).
The P3T3 method computes the NOx emission index (EINOx) of an engine at a specific operation condition as a function of the combustor pressure ($P_3$) and FAR.
The P3T3 method uses the EINOx level at sea-level operating condition as a reference point to scale the desired operating condition EINOx.
Finally, the EINOx level is scaled by a humidity factor based on the water content of the specific operating condition compared to the sea-level water content.
The P3T3 method is give below:

@@

\begin{table}[H]
    \centering
    \caption{Vapor recovery loop water fraction optimization for JetA fuel.}
    \begin{tabular}{|c|c|c|c|c|c|}
        \hline
        Category    & Name          & Value   & Lower & Upper  & Units                 \\
        \hline
        Objective   & $TSEC_{CRZ}$  & 7886.58 & -     & -      & $\frac{BTU}{hr-lb_f}$ \\
        \hline
        Variables   & $x_{H2O,TOC}$ & 0.0699  & 0.0   & -      & -                     \\
                    & $x_{H2O,RTO}$ & 0.0     & 0.0   & -      & -                     \\
                    & $x_{H2O,SLS}$ & 0.0     & 0.0   & -      & -                     \\
                    & $x_{H2O,CRZ}$ & 0.27    & 0.0   & 0.27   & -                     \\
        \hline
        Constraints & $F_{net,TOC}$ & 5800.0  & -     & 5800.0 & $lbf$                 \\
        \hline
    \end{tabular}
    \label{tab_jeta_opt}
\end{table}

\begin{table}[H]
    \centering
    \caption{Vapor recovery loop water fraction optimization for H2 fuel.}
    \begin{tabular}{|c|c|c|c|c|c|}
        \hline
        Category    & Name          & Value   & Lower & Upper  & Units                 \\
        \hline
        Objective   & $TSEC_{CRZ}$  & 7840.04 & -     & -      & $\frac{BTU}{hr-lb_f}$ \\
        \hline
        Variables   & $x_{H2O,TOC}$ & 0.105   & 0.0   & -      & -                     \\
                    & $x_{H2O,RTO}$ & 0.0     & 0.0   & -      & -                     \\
                    & $x_{H2O,SLS}$ & 0.0     & 0.0   & -      & -                     \\
                    & $x_{H2O,CRZ}$ & 0.19    & 0.0   & 0.19   & -                     \\
        \hline
        Constraints & $F_{net,TOC}$ & 5800.0  & -     & 5800.0 & $lbf$                 \\
        \hline
    \end{tabular}
    \label{tab_h2_opt}
\end{table}

@@

\section{Conclusions}
A novel water vapor recovery loop was developed, studied, and optimized for two different types of fuel.
New pyCycle components, a water injector and water extractor, were developed for this project and implmented in the NASA N3 ultra-high bypass turbofan engine.
Additionally, nitrogen oxide emission models were introduced and implemented to allow for NOx emission constraints on the engine model.
The parametric studies that were performed show a strong coupling between the on-design and off-design operation conditions of the N3 engine.
Furthermore, the simple optimizations that were performed for each fuel show an improvement in performance with water injection and recovery.
The different fuels ended up having very similar thrust-specific energy consumption metrics with a slight improvement when using hydrogen as a fuel as opposed to the traditional JetA fuel.
These results are promising as they indicate that the efficiency improvement that has been seen in empirical results can be modeled in zero-dimensional modeling software packages such as pyCycle, allowing multidisciplinary design and analysis for future high-efficiency engine and aircraft design.
Future plans for this project are to use this model to run large parallel sweeps of the model design space and to eventually create a surrogate model of this engine for use in other optimization libraries such as OpenConcept.
Additionally, water condensor models may be researched in the future so that a fairly accurate condensor and pressure loss model can be implemented for the extractor to determine how severe the pressure loss penalties are.
Finally, a pareto front will be computed to determine the tradeoffs between TSEC and NOx emissions for the JetA fuel.

>>>>

\RequirePackage{fix-cm}
% \documentclass[12pt]{article}
\documentclass[12pt]{new-aiaa}

@@

Water recirculation is the process of extracting water from the exhaust stream of a propulsion system and injecting it upstream of the combustor as finely atomized droplets.
NASA, Boeing, and Rolls-Royce studied injectiong water into the airstream and suggested that this technique reduces the NOx emissions as much as 47 percent~\cite{nasa_inject}.
Additionally, water recirculation improves fuel efficiency and thrust output with lower burner temperatures that can improve the lifetime of turbine blades and reduce noise~\cite{nasa_inject}.
Traditional propulsion systems that burn hydrocarbon fuels would require external water storage on the aircraft because they do not produce enough in the exhaust for recirculation.
The added weight of tanks, pumping, and ducting makes this concept infeasible for a conventional aircraft.
However, if hydrogen fuels are used, the main product of hydrogen combustion is water vapor and can thus be recovered.
The ability to recirculate water vapor from the exhaust of hydrogen combustion reduces the requirement for storage tanks and allows for the creation of a closed loop system inside the propulsion cycle.
Additionally, Pratt and Whitney are currently looking at closed-loop water vapor recovery for impementation in their Hydrogen Steam-Injected, Inter‐Cooled Turbine Engine (HySIITE) as part of the US Departement of Energy's Advanced Research Projects Agency-Energy (ARPA-E) program \cite{hysiite}.
Pratt and Whitney suggest that this type of hydrogen-powered engine could achieve zero-carbon emissions and reduce NOx emissions by as much as 80\%.

@@

\noindent
To model this complicated water recirculation loop in pyCycle new components had to be developed.
The two new components that were developed are a water injector to add water to the flow and a water extractor to divert a fraction of the water in the flow back upstream.
The injector component is based on the combustor component already included in pyCycle which operates by injecting a given fuel-to-air ratio (FAR) or fuel flow rate to determine the mass of fuel added to the incoming flow.
This new mixture combination is then used to compute the various chemical species present in the flow at the given thermodynamic state which is determined by the incoming flow.
The new species composition and thermodynamic variables are then determined using a Gibbs Free Energy minization approach called Chemical Equilibrium with Applications (CEA).
Once the new composition is determined, the various thermodynamic variables are passed to the next component in the engine.
The water injector component was created to work similarly to the combustor but would instead inject water as a reactant instead of the fuel.
The water injector can take either water-to-air ratio (WAR) or water mass flow rate to determine how much water is added to the flow.
A simple schematic of the injector is shown in Figure \ref{fig:injector} where $Y_{H_2O}$ is the mole fraction of water molecules.

@@

\noindent
The water extractor component is based on an air-bleed component in pyCycle with the added complexity of extracting a fraction of a specific species within the flow.
This extractor component works by first determining the water content of the incoming flow and extracting a given fraction of that water flow.
The water content of the incoming flow is determined by solving a CEA analysis of the incoming flow.
This water content in the incoming flow is then used with a given fraction of the water to extract to determine the how much mass flow to remove from the stream.
At the some time, the atomic mixture is updated to represent the mixture that would be left if a number of hydrogen and oxygen atoms corresponding to the amount of water were removed.
The outputs of the extractor component are the flow rate of the main flow and the corresponding properties in addition to a water mass flow rate.
While this process certainly would have pressure loss penalties in terms of condensing water in a condersor, this project looks to just look at the effects of water recovery since there are no current models for nozzle exhaust condensors.
A simple schematic of the extractor is shown in Figure \ref{fig:extractor} where $Y_{H_2O}$ is the mole fraction of water molecules and $X_{H_2O,k}$ is the fraction of water that is recovered from the core stream of operating condition, $k$.

@@

\subsection{NOx Model}
To compute the Nitrogen Oxide (NOx) emissions from the engine model when using JetA as a fuel, the P3T3 NOx correlation was used \cite{Voet2021}.
The P3T3 method computes the NOx emission index (EINOx) of an engine at a specific operation condition as a function of the combustor pressure ($P_3$) and FAR.
The P3T3 method uses the EINOx level at sea-level operating condition as a reference point to scale the desired operating condition EINOx.
Finally, the EINOx level is scaled by a humidity factor based on the water content of the specific operating condition compared to the sea-level water content.
The P3T3 method is give below:

@@

\begin{table}[H]
    \centering
    \caption{Vapor recovery loop water fraction optimization for JetA fuel.}
    \begin{tabular}{|c|c|c|c|c|c|}
        \hline
        Category    & Name          & Value   & Lower  & Upper & Units                 \\
        \hline
        Objective   & $TSEC_{CRZ}$  & 7886.58 & -      & -     & $\frac{BTU}{hr-lb_f}$ \\
        \hline
        Variables   & $x_{H2O,TOC}$ & 0.0699  & 0.0    & -     & -                     \\
                    & $x_{H2O,RTO}$ & 0.0     & 0.0    & -     & -                     \\
                    & $x_{H2O,SLS}$ & 0.0     & 0.0    & -     & -                     \\
                    & $x_{H2O,CRZ}$ & 0.27    & 0.0    & 0.27  & -                     \\
        \hline
        Constraints & $F_{net,TOC}$ & 5800.0  & 5800.0 & -     & $lbf$                 \\
        \hline
    \end{tabular}
    \label{tab_jeta_opt}
\end{table}

\begin{table}[H]
    \centering
    \caption{Vapor recovery loop water fraction optimization for H2 fuel.}
    \begin{tabular}{|c|c|c|c|c|c|}
        \hline
        Category    & Name          & Value   & Lower  & Upper & Units                 \\
        \hline
        Objective   & $TSEC_{CRZ}$  & 7840.04 & -      & -     & $\frac{BTU}{hr-lb_f}$ \\
        \hline
        Variables   & $x_{H2O,TOC}$ & 0.105   & 0.0    & -     & -                     \\
                    & $x_{H2O,RTO}$ & 0.0     & 0.0    & -     & -                     \\
                    & $x_{H2O,SLS}$ & 0.0     & 0.0    & -     & -                     \\
                    & $x_{H2O,CRZ}$ & 0.19    & 0.0    & 0.19  & -                     \\
        \hline
        Constraints & $F_{net,TOC}$ & 5800.0  & 5800.0 & -     & $lbf$                 \\
        \hline
    \end{tabular}
    \label{tab_h2_opt}
\end{table}

@@

\section{Conclusions}
A novel water vapor recovery loop was developed, studied, and optimized for two different types of fuel.
New pyCycle components, a water injector and water extractor, were developed for this project and implmented in the NASA N3 ultra-high bypass turbofan engine.
Additionally, nitrogen oxide emission models were introduced and implemented to allow for NOx emission constraints on the engine model.
The parametric studies that were performed show a strong coupling between the on-design and off-design operation conditions of the N3 engine.
Furthermore, the simple optimizations that were performed for each fuel show an improvement in performance with water injection and recovery.
The different fuels ended up having very similar thrust-specific energy consumption metrics with a slight improvement when using hydrogen as a fuel as opposed to the traditional JetA fuel.
These results are promising as they indicate that the efficiency improvement that has been seen in empirical results can be modeled in zero-dimensional modeling software packages such as pyCycle, allowing multidisciplinary design and analysis for future high-efficiency engine and aircraft design.
Future plans for this project are to use this model to run large parallel sweeps of the model design space and to eventually create a surrogate model of this engine for use in other optimization libraries such as OpenConcept.
Additionally, water condensor models may be researched in the future so that a fairly accurate condensor and pressure loss model can be implemented for the extractor to determine how severe the pressure loss penalties are.
Finally, a pareto front will be computed to determine the tradeoffs between TSEC and NOx emissions for the JetA fuel.
All of these future studies will help determine if the claims of performance enhancements and emission reductions of water recovery and hydrogen engines that are promoted by industy are within the realm of possibliy.
